Pattern matching for\\ $(231,213)$-avoiding permutations

### 2015-02-01 article.tex
introduction

@@

A \emph{permutation} of length $n$ is a one-to-one function from an
$n$-element set to itself.
We write permutations as words 
$\sigma = \sigma_1\,\sigma_2\,\ldots\,\sigma_n$, whose letters are distinct
and uasually consist of the integers $1\,2\,\ldots\,n$.
For the sake of convenience, 
we let 
$\pi[i:j]$ stand for 
$\pi_i\,\pi_{i+1}\,\ldots\,\pi_j$,
$\pi[:j]$ stand for $\pi[1:j]$ and
$\pi[i:]$ stand for $\pi[i:n]$.
We let $S_n$ denote the set of all permutations of length $n$.

@@

A permutation $\pi$ is said to \emph{contain} the permutation $\sigma$ 
if there exists a subsequence of (not necessarily consecutive) 
entries of $\pi$ that has the same relative order as $\sigma$, 
and in this case $\sigma$ is said to be a \emph{pattern} of 
$\pi$, written $\sigma \preceq \pi$. 
Otherwise, $\pi$ is said to \emph{avoid} the permutation $\sigma$. 
For example, the permutation $\pi = 3\,9\,1\,8\,6\,7\,4\,5\,2$ 
contains the pattern $\sigma = 51342$, 
as can be seen in the highlighted subsequence of 
$\pi = 3\,\mathbf{9}\,\mathbf{1}\,8\,\mathbf{6}\,\mathbf{7}\,\mathbf{4}\,5\,2$
(or 
$\pi = 3\,\mathbf{9}\,\mathbf{1}\,8\,\mathbf{6}\,\mathbf{7}\,4\,\mathbf{5}\,2$
or
$\pi = 3\,\mathbf{9}\,\mathbf{1}\,8\,\mathbf{6}\,\mathbf{7}\,4\,5\,\mathbf{2}$
). 
Each subsequence $(9\,1\,6\,7\,4$, 
$9\,1\,6\,7\,5$, 
$9\,1\,6\,7\,2$) is called a 
\emph{copy}, 
\emph{instance}, or 
\emph{occurrence} of $\sigma$. 
Since the permutation $\pi = 3\,9\,1\,8\,6\,7\,4\,5\,2$  contains no increasing subsequence of length four, $\pi$ avoids $1\,2\,3\,4$.

Suppose $P$ is a set of permutations. We let $\AV_n(p)$ denote the
set of all $n$-permutations avoiding each permutation in $P$.
For the sake of convenience, we omit $P$'s braces thus having
e.g. $\AV_3(132,4312)$ instead of
$\AV_3(\{132,4312\})$.
A basic example is if 
$\pi = \pi_1\,\pi_2\,\ldots\,\pi_n$ avoids $3\,2\,1$, i.e., 
has no decreasing subsequence of length $3$, then its reverse, 
$\pi' = \pi_n\,\pi_{n-1}\,\ldots\,\pi_1$ avoids $1\,2\,3$, i.e., 
has no increasing subsequence of length $3$.

An \emph{ascent} of a permutation $\pi \in S_n$ is any position 
$1 \leq i < n$ where the following value is bigger than the current one. 
That is, if $\pi = \pi_1\,\pi_2\,\ldots\,\pi_n$, then
$i$ is an ascent if $\pi_i < \pi_{i+1}$.
For example, the permutation 
$3\,4\,5\,2\,1\,6\,$7 has ascents (at positions) $1,2,5,6$.
Similarly, a \emph{descent} is a position 
$1 \leq i < n$ with $\pi_i > \pi_{i+1}$, 
so every $i$ with $1 \leq i < n$ either is an ascent or is a descent of 
$\pi$.
Let $\ustep$ and $\dstep$ denote an ascend and a descend, respectively.
The \emph{stripe} $s_\pi$ of a permutation $\pi \in S_n$ is the word
$\stripe{\pi}{1} \stripe{\pi}{2} \ldots \stripe{\pi}{n-1} \in \{\ustep,\dstep\}^{n-1}$ 
defined by 
$ \stripe{\pi}{i}= \ustep$ if $i$ an ascent in $\pi$ and
$\stripe{\pi}{i} = \dstep$ if $i$ a descent in $\pi$.
For example the stripe of the permutation 
$\pi = 12\,11\,10\,1\,2\,3,4\,9\,8\,5\,6\,7$ 
is 
$\stripew{\pi} = \dstep\dstep\dstep\ustep\ustep\ustep\ustep\dstep\dstep\ustep\ustep$.
The stripe $s\stripew{\sigma}$ is a \emph{pattern} of the strip $\stripew{\pi}$
(or $s\stripew{\sigma}$ \emph{occurs} of the strip $\stripew{\pi}$) if $\stripew{\sigma}$ occurs in
$\stripew{\pi}$ as a subsequence.

@@

A \emph{bivincular pattern} (abbreviated BVP) $\sigma$
of length $k$ is a permutation in $S_k$ written in
two-line notation 
(that is the top row is $1\,2\,\ldots\,k$ and the bottom row
is a permutation $\sigma_1\,\sigma_2\,\ldots\,\sigma_k$).
We have the following conditions on the top and bottom rows
of $\sigma$:
\begin{itemize}
	\item
	If the bottom line of $\sigma$ contains
	$\underline{\sigma_i\,\sigma_{i+1}\,\ldots\,\sigma_j}$
	then the letters corresponding to 
	$\sigma_i\,\sigma_{i+1}\,\ldots\,\sigma_j$ in an occurrence of
	$\sigma$ in a permutaton must be adjacent, whereas there is
	no adjacency condition for
	non-underlined consecutive letters.
	Moreover if the bottom row of $\sigma$ begins with
	$_\llcorner{\sigma_1}$ then any occurrence of $\sigma$ 
	in a permutation $\pi$ must begin with the leftmost 
	letter of $\pi$,
	and 
	if the bottom row of $\sigma$ begins with
	${\sigma_k}_\lrcorner$ then any occurrence of $\sigma$ 
	in a permutation $\pi$ must end with the rightmost 
	letter of $\pi$.
	\item
	If the top line of $\sigma$ contains
	$\overline{i\,i+1\,\ldots\,j}$ then the letters corresponding to
	$\sigma_i, \sigma_{i+1}, \ldots, \sigma_j$ in an
	occurrence of $\sigma$ in a permutation must be adjacent in values,
	whereas there is no value adjacency restriction for non-overlined 
	letters.
	Moreover, if the top row of $\sigma$ begins with
	$^ulcorner{1}$ then
	any occurrence of $\sigma$ is a permutation $\pi$ must begin with
	the smallest letter of $\pi$, and
	if top row of $\sigma$ ends with $k^\urcorner$ then
	any occurrence of $\sigma$ is a permutation $\pi$ must end with
	the largest letter of $\pi$.
\end{itemize}

@@

		\textbf{Lemma  \numl{firstelement} \thenum} If a permutation is in $\Av{n}{213,231}$ 
		then the first element is either
		the minimal or the maximal element.\\

		\textbf{Proof} Suppose that we have
		$\pi \in \mathcal{S}_{n}$, and
		$\pi(1) \neq 1 $ and $\pi(1) \neq n$ then we have either
		$\pi(1) ... n ... 1$ or $\pi(1) ... 1 ... n $, either the motif 231 or 213, which is not possible by hypothesis. $\Box$\\

		\textbf{Corollary  \num \thenum} A permutation $\pi \in \Perm{n}$ is in $\Av{n}{213,231}$ if and only if for $1 \leq i \leq n$, 
		$\pi[i:]$ start either with the maximal element
		or the minimal element of $\pi[i:]$.\\

		\textbf{Corollary \numl{stripe} \thenum}  If $\pi \in \Av{n}{213,231}$ its stripe can be define such as, \\
		$\stripe{\pi}{i} = \left\{ 
  		\begin{array}{l l}
   			\ustep & \quad \text{if $\pi(i)$ is the minimal element of $\pi[i:]$}\\    
   			\dstep & \quad \text{if $\pi(i)$ is the maximal element of $\pi[i:]$}\\
	 	\end{array} \right. 
	 	$\\

@@

		\textbf{Proof} By induction on the size of the permutation :
		the proposition is true for the permutations $(1,2)$ and $(2,1)$.
		Suppose the proposition true for $\Av{n}{213,231}$.
		Let $\pi \in \Av{n}{213,231}$, 
		we show that the proposition holds for $\pi$.
		Let $\pi'$ be the permutation $\pi[2:]$.
		$\pi' \in \Av{n}{213,231}$, so
		every element of $p'$ verify our lemma (by induction).
		We still need to compare the first element with the rest.
		But the first element
		is the maximal (minimal) element of all the permutation 
		(by Lemma \ref{firstelement}).
		Moreover if the first element is the maximal (minimal)
		it will mapped to an descent (ascent).
		So the proposition is also true
		for the first element.
		$\Box$\\

>>>>

	We focus in this paper on $(213,231)$-avoiding permutation. 
	The number of $n$-permutations that avoid both 
	$231$ and $213$ is 
	$t_0 = 1$ for $n = 0$ and 
	$t_n =2^{n−1}$ for $n\geq 1$ \cite{Simion:Schmidt:EJC:1985}.
	The permutations that do not contain the permutation pattern $231$
	are exactly the \emph{stack-sortable permutations} and they are counted by 
	the Catalan numbers \cite{Knuth:1997:ACP:260999}.
	A stack-sortable permutation is a permutation whose elements may be sorted by 
	an algorithm whose internal storage is limited to a single stack data structure.

@@

A \emph{permutation} of length $n$ is a one-to-one function from an
$n$-element set to itself.
We write permutations as words 
$\sigma = \sigma_1\sigma_2\,\ldots\,\sigma_n$, whose letters are distinct
and uasually consist of the integers $12\,\ldots\,n$.
For the sake of convenience, 
we let 
$\pi[i:j]$ stand for 
$\pi_i\pi_{i+1}\,\ldots\,\pi_j$,
$\pi[:j]$ stand for $\pi[1:j]$ and
$\pi[i:]$ stand for $\pi[i:n]$.
We let $S_n$ denote the set of all permutations of length $n$.

@@

A permutation $\pi$ is said to \emph{contain} the permutation $\sigma$ 
if there exists a subsequence of (not necessarily consecutive) 
entries of $\pi$ that has the same relative order as $\sigma$, 
and in this case $\sigma$ is said to be a \emph{pattern} of 
$\pi$, written $\sigma \preceq \pi$. 
Otherwise, $\pi$ is said to \emph{avoid} the permutation $\sigma$. 
For example, the permutation $\pi = 391867452$ 
contains the pattern $\sigma = 51342$, 
as can be seen in the highlighted subsequence of 
$\pi = 3\mathbf{9}\mathbf{1}8\mathbf{6}\mathbf{7}\mathbf{4}52$
(or 
$\pi = 3\mathbf{9}\mathbf{1}8\mathbf{6}\mathbf{7}4\mathbf{5}2$
or
$\pi = 3\mathbf{9}\mathbf{1}8\mathbf{6}\mathbf{7}45\mathbf{2}$
). 
Each subsequence $(91674$, 
$91675$, 
$91672$) is called a 
\emph{copy}, 
\emph{instance}, or 
\emph{occurrence} of $\sigma$. 
Since the permutation $\pi = 391867452$  contains no increasing subsequence of length four, $\pi$ avoids $1234$.

Suppose $P$ is a set of permutations. We let $\AV_n(p)$ denote the
set of all $n$-permutations avoiding each permutation in $P$.
For the sake of convenience, we omit $P$'s braces thus having
e.g. $\AV_3(132,4312)$ instead of
$\AV_3(\{132,4312\})$.
A basic example is if 
$\pi = \pi_1\pi_2\,\ldots\,\pi_n$ avoids $321$, i.e., 
has no decreasing subsequence of length $3$, then its reverse, 
$\pi' = \pi_n\pi_{n-1}\,\ldots\,\pi_1$ avoids $123$, i.e., 
has no increasing subsequence of length $3$.

An \emph{ascent} of a permutation $\pi \in S_n$ is any position 
$1 \leq i < n$ where the following value is bigger than the current one. 
That is, if $\pi = \pi_1\pi_2\,\ldots\,\pi_n$, then
$i$ is an ascent if $\pi_i < \pi_{i+1}$.
For example, the permutation 
$345216$7 has ascents (at positions) $1,2,5,6$.
Similarly, a \emph{descent} is a position 
$1 \leq i < n$ with $\pi_i > \pi_{i+1}$, 
so every $i$ with $1 \leq i < n$ either is an ascent or is a descent of 
$\pi$.
Let $\ustep$ and $\dstep$ denote an ascend and a descend, respectively.
The \emph{stripe} $s_\pi$ of a permutation $\pi \in S_n$ is the word
$\stripe{\pi}{1} \stripe{\pi}{2} \ldots \stripe{\pi}{n-1} \in \{\ustep,\dstep\}^{n-1}$ 
defined by 
$ \stripe{\pi}{i}= \ustep$ if $i$ an ascent in $\pi$ and
$\stripe{\pi}{i} = \dstep$ if $i$ a descent in $\pi$.
For example the stripe of the permutation 
$\pi = 12\,11\,10\,1\,2\,3,4\,9\,8\,5\,6\,7$ 
is 
$\stripew{\pi} = \dstep\dstep\dstep\ustep\ustep\ustep\ustep\dstep\dstep\ustep\ustep$.
The stripe $s\stripew{\sigma}$ is a \emph{pattern} of the strip $\stripew{\pi}$
(or $s\stripew{\sigma}$ \emph{occurs} of the strip $\stripew{\pi}$) if $\stripew{\sigma}$ occurs in
$\stripew{\pi}$ as a subsequence.

@@

A \emph{bivincular pattern} (abbreviated BVP) $\sigma$
of length $k$ is a permutation in $S_k$ written in
two-line notation 
(that is the top row is $12\,\ldots\,k$ and the bottom row
is a permutation $\sigma_1\sigma_2\,\ldots\,\sigma_k$).
We have the following conditions on the top and bottom rows
of $\sigma$:
\begin{itemize}
	\item
	If the bottom line of $\sigma$ contains
	$\underline{\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j}$
	then the letters corresponding to 
	$\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j$ in an occurrence of
	$\sigma$ in a permutaton must be adjacent, whereas there is
	no adjacency condition for
	non-underlined consecutive letters.
	Moreover if the bottom row of $\sigma$ begins with
	$_\llcorner{\sigma_1}$ then any occurrence of $\sigma$ 
	in a permutation $\pi$ must begin with the leftmost 
	letter of $\pi$,
	and 
	if the bottom row of $\sigma$ begins with
	${\sigma_k}_\lrcorner$ then any occurrence of $\sigma$ 
	in a permutation $\pi$ must end with the rightmost 
	letter of $\pi$.
	\item
	If the top line of $\sigma$ contains
	$\overline{i\,i+1\,\ldots\,j}$ then the letters corresponding to
	$\sigma_i, \sigma_{i+1}, \ldots, \sigma_j$ in an
	occurrence of $\sigma$ in a permutation must be adjacent in values,
	whereas there is no value adjacency restriction for non-overlined 
	letters.
	Moreover, if the top row of $\sigma$ begins with
	$^ulcorner{1}$ then
	any occurrence of $\sigma$ is a permutation $\pi$ must begin with
	the smallest letter of $\pi$, and
	if top row of $\sigma$ ends with $k^\urcorner$ then
	any occurrence of $\sigma$ is a permutation $\pi$ must end with
	the largest letter of $\pi$.
\end{itemize}

@@

		\begin{lemma}[Folklore]
		\label{lemma:first element is 1 or n}
		The 􏰅first element of any $(213,231)$-avoiding permutation
		must be either $1$ or $n$.
		\end{lemma}

		\begin{proof} 
		Any other initial element would serve as a `$2$' in either a 
		$231$ or $213$ with $1$ and $n$ as the `$1$' and `$3$' respectively.
		\end{proof}

		The above lemma gains in interest in the form of the following corollaries.

		\begin{corollary}
		$\pi \in \AV_n(213,231)$ if and only if for $1 \leq i \leq n$, 
		$\pi[i:]$ start either with the maximal element
		or the minimal element of $\pi[i:]$.
		\end{corollary}

		\begin{corollary}
		Let $\pi \in \AV_n(213,231)$. We have
		$$
		\stripe{\pi}{i} =
		\begin{cases}
   			\ustep & \text{if $\pi(i)$ is the minimal element of $\pi[i:]$}\\    
   			\dstep & \text{if $\pi(i)$ is the maximal element of $\pi[i:]$}\\
	 	\end{cases}
	 	$$
	 	\end{corollary}

@@

		\textbf{Proof} By induction on the size of the permutation :
		the proposition is true for the permutations $(1,2)$ and $(2,1)$.
		Suppose the proposition true for $\Av{n}{213,231}$.
		Let $\pi \in \Av{n}{213,231}$, 
		we show that the proposition holds for $\pi$.
		Let $\pi'$ be the permutation $\pi[2:]$.
		$\pi' \in \Av{n}{213,231}$, so
		every element of $p'$ verify our lemma (by induction).
		We still need to compare the first element with the rest.
		But the first element
		is the maximal (minimal) element of all the permutation 
		(by Lemma \ref{lemma:first element is 1 or n}).
		Moreover if the first element is the maximal (minimal)
		it will mapped to an descent (ascent).
		So the proposition is also true
		for the first element.
		$\Box$\\
